msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-09 paper.tex

\subsection*{Population structure}
We support all the stuff that ms does.
Demographic events where we change
the properties of the population.

We provide a debugger interface which allows the user to see the
various population sizes and growth rates over time. This includes
a way to compute coalescence rate trajectory, which provides a useful
ground truth when comparing with methods that estimate population
size over time~\citep{adrion2019community}.

The interface
is very flexible and general. We also reading in population tree
descriptions as produced by programs such as
StarBeast~\citep{heled2009bayesian}.

We support tracking lineages through time by associating a population
with each node, and having migration records. We also have
census events, which will record the location of every lineage
at a particular time.

@@

\subsection*{Simulation interface}
The primary interface for \msprime\ is through a Python application
programming interface (API). This has many advantages. One is
that we can grow the interface over time. Another is that
downstream applications such as \texttt{stdpopsim}~\citep{adrion2019community}
can be developed.

@@

\subsection*{Computing coalescence rates and lineage location probabilities}
Many popular approaches in population genetics use the distribution of coalescence rates between
pairs of lineages in one or more populations to infer effective population sizes over time
\cite{li2011inference,sheehan2013estimating,schiffels2014inferring}
or split times
and subsequent migration rates between populations
\cite{wang2020tracking}.
Given a demographic model, the
\texttt{msprime} DemographyDebugger is able to compute coalescence rates for two or more lineages drawn
from one or more populations at specified times in the past, providing a ground truth for coalescent rate-based
inference methods. To compute coalescence rates for a set of times $Ts=(t_0, t_1, \ldots)$ in the
past, we write
\begin{lstlisting}[frame=single]
dbg = demography.debug()
rates = dbg.coalescence_rate_trajectory(Ts, num_samples)
\end{lstlisting}
where \texttt{num\_samples} is a list with length equal to the number of populations that
specifies the number of sampled lineages from each population. For example,
to get pairwise rates within the same
population, this list would contain zeros for all populations except for the population index of interest,
which would have a $2$. To get the rate of coalescence for two lineages sampled from different
populations, we would set \texttt{num\_samples} to have ones in those population indices.

More simply, \texttt{msprime} can also compute the probability of a lineage that was in a given
population at some time to be found in other populations (through continuous or mass migration events).
For a given sample time $t\geq0$ and any list of times $Ts$ in the past, we write
\begin{lstlisting}[frame=single]
prob = dbg.lineage_probabilities(Ts, sample_time=t)
\end{lstlisting}
This function is also used to determine the time intervals that lineages can be found in each population,
given a list of contemporary and/or ancient samples. For a given sampling configuration, by writing
\begin{lstlisting}[frame=single]
lineage_dict = dbg.possible_lineage_locations(samples=samples)
\end{lstlisting}
we get a dictionary with epoch intervals as keys whose values are a list that indicates True or False for
whether each population could contain ancestral material to our samples.
If no sampling configuration is provided, we assume we sample lineages from each population at time zero.

@@

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
opens the possibility for new applications of coalescent simulations.
Also \stdpopsim\ is cool~\citep{adrion2019community}

>>>>

\subsection*{Demography}
% TODO should we put in some citations here for the models? Hardly seems
% necessary.
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are provided, which allow for
varying population parameters and migration rates over time, as well as
population splits, admixtures and pulse migrations. The location of lineages
over time can be tracked in as much detail as required: each tree node is
automatically associated with the population in which it arose, the location of
lineages can be recorded at any given time via census events, or every lineage
migration can be recorded.

A major change for \msprime\ 1.0 is the introduction of the new Demography API,
designed to address a serious flaw in the \msprime\ 0.x interface which lead to
a number of avoidable errors in downstream
simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
separate parameters be provided to the \texttt{simulate} function to describe a
demographic model, making it easy to accidentally omit information. The 1.0 API
resolves this issue by creating a new \texttt{Demography} class, which
encapsulates all information about the demographic model, and fully decouples
the definition from other simulation details. An instance of this class is then
provided as a parameter to the new \texttt{sim\_ancestry} function,
substantially reducing the potential for error. Another improvement over the
0.X APIs is the introduction of explicit population split and admixture events,
and a population state machine that ensures that lineages cannot migrate into
(or be sampled from) inactive populations. This demography model is compatible
with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
class supports importing Demes models. We also support importing demographic
models in the form of Newick species tree descriptions, as well as those
produced by programs such as *BEAST~\citep{heled2009bayesian}.

Many popular approaches in population genetics use the distribution of
coalescence rates between pairs of lineages in one or more populations to infer
effective population sizes over
time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
split times and subsequent migration rates between
populations~\citep{wang2020tracking}. The \texttt{DemographyDebugger} provides
methods to compute a suite of numerical predictions about the location of
lineages over time, for example the coalescence rates for two or more lineages
drawn from populations at specified times in the past. This provides a valuable
ground-truth when evaluating such inference methods, as illustrated by
\cite{adrion2020community}.

@@

\subsection*{Simulation interface}
The primary interface for \msprime\ is through a Python application
programming interface (API). This has many advantages. One is
that we can grow the interface over time. Another is that
downstream applications such as \texttt{stdpopsim}~\citep{adrion2020community}
can be developed.

The \texttt{msp simulate} command also supports demographic models
specified in Demes format~\citep{gower2021demes}, allowing users with
no Python experience to run complex demographic simulations.

@@

@@

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
opens the possibility for new applications of coalescent simulations.
Also \stdpopsim\ is cool~\citep{adrion2020community}
